---
title: "Predict the President (Neural Networks)"
author: "Kenneth Ssekimpi"
student_number: "SSKKEN001"
assignment: "Assignment 1"
format: html
editor: visual
embed-resources: true
fonts: Times New Roman
bibliography: references.bib
execute:
  echo: false
  warning: false
---

```{r setup, include=FALSE}

knitr::opts_knit$set(root.dir = "C:/Users/User/OneDrive/Documents/School/2023/Masters/STA5073Z/Assignments/Assignment 1/ds4l-assignment-1/")

knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
```

```{r libraries}
# Clear global environment
# rm(list=ls())

# Libraries we need
libs <- c("caret", 'dplyr', "gbm", 'ggplot2', "kableExtra", 'keras', 'lubridate',  'quarto', "randomForest", 'readr', "rpart", 'stringr', "tensorflow", 'tfhub', 'tidytext', 'tm', 'topicmodels', 'wordcloud')

# Install missing libraries
installed_libs <- libs %in% rownames(installed.packages())
if (any(installed_libs == FALSE)) {
  install.packages(libs[!installed_libs], repos='http://cran.us.r-project.org')
}

# Load libraries
invisible(lapply(libs, library, character.only = TRUE))

use_virtualenv("myenv", required = TRUE) 
```

```{r}
unzip("sona-addresses-1994-2023.zip", exdir = "data")
```

# Abstract

# Introduction

The State of the Nation Address (SONA) of the President of South Africa is an annual event in which the President reports on the status of the nation to a joint sitting of Parliament. SONA speeches are an important source of information on the South African government's priorities, challenges, and plans.

In this paper, we conduct a

Our work contributes to the understanding of the South African government's priorities, challenges, and plans over time. It also provides insights into the public discourse on these issues.

```{r read_wrangle}
# Get a list of all text files in the directory
text_files <- list.files(path = "data", pattern = ".txt")

# Initialize an empty list to store the data
speech_data <- c()
i = 0
num_chars <- c(27050, 12786, 39019, 39524, 37489, 45247, 34674, 41225, 37552, 41719, 50544, 58284, 34590, 39232, 54635, 48643, 48641, 44907, 31101, 47157, 26384, 33281, 33376, 36006, 29403, 36233, 32860, 32464, 35981, 33290, 42112, 56960, 47910, 43352, 52972, 60000)

# Loop through the list of text files and read them into R
for (file in text_files) {
  i = i + 1
  # speech <- readLines(file, warn = FALSE)
  # Open the file for reading
  file_handle <- file(paste("data/", file, sep = ""), "r")
  speech <- readChar(file_handle, nchars = 60000)
  # speech_data[[file]] <- speech
  speech_data[i] <- speech
  # Close the file
  close(file_handle)
}

sona <- data.frame(filename = text_files, speech = speech_data, stringsAsFactors = FALSE)

# extract year and president for each speech
sona$year <- str_sub(sona$filename, start = 1, end = 4)
sona$president <- str_remove_all(str_extract(sona$filename, "[dA-Z].*\\."), "\\.")

# clean the sona dataset by adding the date and removing unnecessary text
replace_reg <- '(http.*?(\\s|.$))|(www.*?(\\s|.$))|&amp;|&lt;|&gt;|\n'

sona <-sona %>%
  mutate(speech = str_replace_all(speech, replace_reg , ' ')
         ,date = str_sub(speech, start=1, end=30)
         ,date = str_replace_all(date, "February", "02")
         ,date = str_replace_all(date, "June", "06")
         ,date = str_replace_all(date, "Feb", "02")
         ,date = str_replace_all(date, "May", "05")
         ,date = str_replace_all(date, "Jun", "06")
         ,date = str_replace_all(date, "Thursday, ","")
         ,date = str_replace_all(date, ' ', '-')        
         ,date = str_replace_all(date, "[A-z]",'')
         ,date = str_replace_all(date, '-----', '')
         ,date = str_replace_all(date, '----', '')
         ,date = str_replace_all(date, '---', '')
         ,date = str_replace_all(date, '--', '')
  )

sona$date[36] <- "09-02-2023"
sona$year[36] <- "2023"


```

```{r preprocess}

sona$date <- dmy(sona$date)
# Assign labels as integers
sona$prez_encoded <- as.integer(factor(sona$president))-1

sona <- sona %>%
  select(date, speech, year, president, prez_encoded)

ggplot(sona, aes(x = date, fill = president)) +
  geom_histogram(position = "identity", bins = 20, show.legend = FALSE)

```

```{r Bag_of_words}

unnest_reg <- "[^A-Za-z_\\d#@']"
replace_reg <- '(https?:.*?(\\s|.$))|(www.*?(\\s|.$))|&amp;|&lt;|&gt;'

tidy_sentences <- sona %>%
  mutate(speech, speech = str_replace_all(speech, "’", "'")) %>%  
  mutate(speech = str_replace_all(speech, replace_reg, '')) %>%  
  unnest_tokens(sentence, speech, token = 'sentences') %>%
  filter(str_detect(sentence, '[a-z]')) %>%
  mutate(Sentence_ID =  row_number())

set.seed(1991)

# Create a document-term matrix (DTM) and convert it to a TF-IDF representation
dtm <- DocumentTermMatrix(Corpus(VectorSource(sona$word)))
tfidf <- weightTfIdf(dtm)
tfidf_matrix <- as.matrix(tfidf)
```

```{r data_split}
# Split the dataset into three subsets: training, validation, and test
set.seed(1991)
sample_size <- floor(0.8 * nrow(sona))
train_indices <- sample(seq_len(nrow(sona)), size = sample_size)

x_train <- tfidf_matrix[train_indices, ]
x_test <- tfidf_matrix[-train_indices, ]

y_train <- to_categorical(sona$leader[train_indices], num_classes = max(sona$leader)+1)
y_test <- to_categorical(sona$leader[-train_indices], num_classes = max(sona$leader)+1)

training_indices <- sample(1:nrow(sona), 0.8 * nrow(sona), replace = FALSE)
validation_indices <- setdiff(1:nrow(sona), training_indices)
test_indices <- validation_indices[sample(1:length(validation_indices), 0.5 * length(validation_indices), replace = FALSE)]

# Create the training, validation, and test datasets
training_set <- sona[training_indices, ]
validation_set <- sona[validation_indices, ]
test_set <- sona[test_indices, ]
```

```{r}



# # Calculate the accuracy of the classifier on the validation set
# validation_accuracy <- mean(predicted_labels == validation_set$president)
# 
# # Calculate the precision of the classifier on the validation set
# validation_precision <- mean(predicted_labels[predicted_labels == 1] == validation_set$president[predicted_labels == 1])
# 
# # Calculate the recall of the classifier on the validation set
# validation_recall <- mean(predicted_labels[predicted_labels == validation_set$president] == 1)
# 
# # Calculate the F1 score of the classifier on the validation set
# validation_f1_score <- 2 * (validation_precision * validation_recall) / (validation_precision + validation_recall)


```

# Predict the President (Neural Networks)

## Abstract

This is a concise summary of our paper...

## Keywords

## Introduction

This section provides more detailed background information on our topic...

## Literature review

...

## Data and methods

This section describes how we conducted our study...

## Results

This sections presents the results of our study...

## Discussion

This section interprets our results and discusses their implications...

## Conclusion

This section summarizes the main findings of our study and states the conclusions that can be drawn from the data...

## References

This section lists all of the sources that we cited in our paper...
